### 2020-06-24 chapters/background.tex
Marco feedback

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine by grouping the results by keyword and counting how many times a keyword was searched for trend analysis purposes. Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop \cite{borthakur2007hadoop} \cite{Hadoop} \cite{ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a distributed computation framework or programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this they read and write from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark} \cite{Spark} was released.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

@@

In Section 2.1 we have defined a data stream as an unbounded data set. But what exactly are the issues of working with an infinite stream of data? Why might we need to convert it into a bounded data set and how to do so? In this Section, we will analyze such issues and present streaming windows as the solution.

@@

For the scope of this Thesis, we will work under a tuple-based true sliding window framework. However, as explained in Section \ref{sec:ft-monitoring}, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, we can obtain a bounded data set from a data stream by applying windowing techniques. When doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. Tangwongsan \emph{et. al} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}. In this example, we wanted to compute the average temperature per rack in the data cencer in the last 5 minutes. If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:

@@

EMAs are recursive: the EMA value at timestamp \textit{t} depends only on the EMA value at timestamp \textit{t-1}. This recursive property allows us to store only one EMA value and continuously update it for each new value val\textsubscript{t}, based on the following recursion formula:

\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda\delta\textsubscript{t}$ + val\textsubscript{t}
\label{def:ema}
\end{definition}

with $\lambda$ as a constant factor applied to the previous EMA value and $\delta\textsubscript{t}$ as the time passed between \textit{t} and \textit{t-1}. This $\lambda$ constant (0 < $\lambda \leq$ 1) is known as the smoothing coefficient. The smoothing or decay factor $\lambda$ is given by:
\begin{equation}
    \lambda = 2^{- \dfrac{1}{\tau}}
    \label{eq:ema-decay}
\end{equation}
where $\tau$ is the half-life of the EMA. The half-life $\tau$ controls how fast old events are forgotten as the smoothing factor depends on it. The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half. For instance, with $\tau = 1s$ we have $\lambda = 0.5$, which means for each 1s step we reduce the weights applied to each event by half.

Note that it is also possible to express a tuple-based half-life, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half. In this case, the definition does not take into account the time passed between events:
\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $\lambda$ + val\textsubscript{t}
\label{def:tuple-ema}
\end{definition}
For example, with a tuple-based half-life of $\tau = 1$ tuple, the smoothing factor is $\lambda = 0.5$, meaning for each new tuple all weights are reduced by half.

>>>>

Batch processing is the processing of data grouped by batches. A batch is a collection of data points that have been grouped by certain criteria. Examples of such criteria could be \textit{"all the transactions from the past two months"}, \textit{"the last million transactions"} or even \textit{"all the transactions made by an user until now"}. Batch data sets are known as static data \cite{Martin-Batch-Defin}. An example application of such a technique would be processing all the keywords searched for in the google search engine, and then grouping the results by keyword to count how many times a keyword was used in searches (for trend analysis purposes). Batch processing is the oldest paradigm between batch and stream processing. Historically speaking, Apache Hadoop \cite{borthakur2007hadoop} \cite{Hadoop} \cite{ApacheHadoop} was one of the first widely known frameworks for large-scale batch processing, created in 2005. Hadoop uses a distributed computation framework or programming model called MapReduce \cite{MapReduce}. This paradigm applies a map function that processes key/value pairs and generates sets of intermediate ones, ultimately merging all intermediate values
using a reducer function. To do this they read and write from disk on intermediate steps, which makes it inefficient for applications that reuse working sets of data across multiple parallel operations (\textit{e.g.} iterative Machine Learning algorithms). Nearly a decade later, Apache Spark \cite{ApacheSpark} \cite{Spark} was released.
Spark retains the scalability and fault tolerance of MapReduce but introduces Resilient Distributed Datasets (RDDs) \cite{SparkRDDs}. An RDD is a read-only collection of objects partitioned across a set of machines that can be rebuilt if a partition is lost.
Additionally, Spark processes data in random access memory (RAM), while Hadoop MapReduce persists data to the disk after a map or reduce action. Thus, Spark achieves lower processing times.

@@

In Section 2.1 we have defined a data stream as an unbounded data set. But what exactly are the issues of working with an infinite stream of data? Why might we need to convert it into a bounded data set and how to do so? In this Section, we will analyze such issues and present sliding windows as the solution.

@@

For the scope of this Thesis, we will work under a tuple-based true sliding window framework. However, despite focusing on tuple-based true sliding windows, our approach is just as valid for time-based windows and other step values as well.

@@

While some aggregations can be computed on an infinite data set, often we want to compute these over a finite one. As we have seen in Section \ref{sec:windows}, we can obtain a bounded data set from a data stream by applying windowing techniques. When doing so, we obtain a finite data set denoted as a sliding window and compute aggregations over it. Tangwongsan \emph{et. al} define such aggregations over sliding windows as Sliding Window Aggregations (SWAGs) \cite{Tangwongsan-Sliding-Window-Aggregation-Algorithms}. As an example of a SWAG, consider the data center temperature monitoring problem from Section \ref{sec:windows}. In this example, we wanted to compute the average temperature per rack in the data center in the last 5 minutes. If we translate this SWAG into streaming SQL following a similar syntax to the one provided by Apache Flink \cite{ApacheFlink} we would get:

@@

EMAs are recursive: the EMA value at timestamp \textit{t} depends only on the EMA value at timestamp \textit{t-1}. This recursive property allows us to store only one EMA value and continuously update it for each new value val\textsubscript{t}, based on the following recursion formula:
\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $2^{-\frac{\delta_t}{\tau_{1/2}}}$ + val\textsubscript{t}
\label{def:ema}
\end{definition}
where $\delta_t$ is the time passed between \textit{t} and \textit{t-1} and $\tau_{1/2}$ is the time-based half-life of the EMA.

The half-life $\tau_{1/2}$ controls how fast old events are forgotten. The half-life is the amount of time that it takes for an element to see its contribution or weight reduced to half. For instance, with $\tau_{1/2} = 1s$, for each $\delta_t = 1s$ step, we reduce the weights applied to each event by half.

Note that it is also possible to express a tuple-based half-life $n_{1/2}$, \textit{i.e.}, how many tuples or events must the EMA process before a certain element has its weight reduced by half. In this case, the definition does not take into account the time passed between events:
\begin{definition}
EMA\textsubscript{t} = EMA\textsubscript{t-1} * $2^{-\frac{1}{n_{1/2}}}$ + val\textsubscript{t}
\label{def:tuple-ema}
\end{definition}
For example, with a tuple-based half-life of $\tau = 1$ tuple, for each new tuple all weights are reduced by half.

### 2020-06-24 chapters/introduction.tex
Marco feedback

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized in terms of its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former is best known as stream processing, whereas the latter as batch processing.

@@

Feedzai's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained before deployment making assumptions on the statistical distributions of future incoming data. Thus their model performance is linked to the maintenance of these assumptions. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.} Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.} new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift.

\section{Objectives} \label{sec:objectives}
Feedzai's data pattern shift challenge is not unique to them nor their field. Many analytical systems are static once deployed and are configured \textit{a priori} under the assumption that future data flowing through the system will roughly follow the same distribution as previously seen data.
The problem is that data pattern shifts will lead to poor performance by the previously configured system. Monitoring the data patterns themselves and alerting for changes in the underlying distribution would allow users to reconfigure the system before its performance declines.

@@

\section{Hypothesis} \label{sec:hypothesis}
The question this Thesis aims to answer is whether sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint.

>>>>

All of these applications generate large volumes of data over time which poses data storage and processing challenges. This data is characterized in terms of its large volume, high velocity and high variety \cite{Mavragani-GoogleTrends-SLR}, creating large scale data management problems. Additionally, the large amounts of information produced by such systems and use-cases lead to the creation of multiple real-time unbounded data sets, also known as data streams. The information that flows through such a stream can be analyzed on-the-fly or stored for later processing. The former is best known as stream processing, whereas the latter is known as batch processing.

@@

Feedzai's solution for fraud detection consists of a workflow of different components. Incoming events are fed into this workflow and processed accordingly. Workflow components can be sets of user-defined rules and/or Machine Learning (ML) models that output scores reflecting their belief on whether a transaction is fraudulent or not. ML models are configured and trained,  before deployment, often with assumptions on the statistical distributions of future incoming data. Thus their model performance depends on how well those assumptions hold for future data. However, fraud patterns are not static over time, but seasonal --- \textit{e.g.} Black Fridays, holiday shopping and product launches --- and evolutionary --- \textit{e.g.} new fraud attack methods by fraudsters. Due to these shifts in data patterns and distributions, the ML models' performance can deteriorate over periods of time. In the Machine Learning field, this performance decay caused by data pattern shifts is also known as concept drift.

\section{Objectives} \label{sec:objectives}
Feedzai's data pattern shift challenge is not unique to them nor their field. Many analytical systems are static once deployed and are configured \textit{a priori} under the assumption that future data flowing through the system will roughly follow the same distribution as previously seen data.
The problem is that data pattern shifts will lead to poor performance of the previously configured system. Monitoring the data patterns themselves and alerting for changes in the underlying distribution would allow users to reconfigure the system before its performance declines.

@@

\section{Hypothesis} \label{sec:hypothesis}
The question this Thesis aims to answer is whether sliding window aggregations can be used to detect data pattern shifts in data streaming scenarios in the presence of high volumes of high velocity, highly skewed and seasonal data in real-time and with a low memory footprint. In this Thesis, we study some sliding window aggregations that do not keep the sliding window in memory but still give an estimate on the aggregation value.

### 2020-06-24 chapters/my-work.tex
Marco feedback

\section{State of the art outlier detection methods}
Our first theoretical approach was to use one of the outlier detection methods presented in Section \ref{sec:outliers}.

We reviewed subsequence and point outlier detection methods. Using point outlier detection methods, the main idea is to keep track of the rate of point outliers detected. For instance, keep track of how many outliers were alerted in the past hour. The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$. On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier when compared to the reference window.

@@

In Section \ref{sec:sota-swag-algs} we review algorithms to compute sliding window aggregations and in Sections \ref{sec:pds} and \ref{sec:sliding-pds} we discussed special aggregators and respective sliding window implementations.

@@

In Figure \ref{fig:approach2-alert-state}, after further sliding steps of our sliding window, we obtain \textit{Window 2}. In \textit{Window 2}, the mean and standard deviation aggregation values change from 5 and 1 to 12 and 6, respectively. In this case, we would measure the difference between the sliding window aggregations and the reference window ones and if above a certain threshold raise an alert. For example, if we used a threshold of $\alpha$ = 6, we would raise an alert, as at least one sliding aggregation, in this case the mean, would differ of at least $\alpha$ when compared to the reference value.

@@

Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection.  
\textcolor{red}{Marco had a problem with this example, dont understand what he meant}
To illustrate this problem, consider both time-series show in Figure \ref{fig:approach2-timeseries}. Time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}. Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two different time-series.

@@

A data stream is a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are time-dependent variables named features. They can be thought of as columns in a traditional relational database table. Our goal is to detect data pattern shifts for each feature in an streaming fashion. Such shifts will be measured between a reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

@@

We do not need to directly compare the entire window contents of the reference and sliding windows. The latter would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size. Instead, we want to use a  \textit{(a)} fixed cost in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

@@

We devise a two-phase method. The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}). The second phase (detailed in Section \ref{sec:stream-phase}) focus on efficiently updating the sliding window histogram for each feature and periodically performs a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be. The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.

\subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
As stated, we compute histogram aggregations for the reference and sliding windows. The computation of the reference histogram is done once in batch and hence an exact histogram can be computed. However, the sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.

In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they compute sliding window aggregations without actually storing the window contents. Definition \ref{def:tuple-ema} shows how to compute an EMA using its recurrence formula. An EMA state or value depends only on the previous one times an exponential decay $\lambda$. Recall that this decay is given by Equation \ref{eq:ema-decay}:
\[ \lambda = 2^{- \dfrac{1}{\tau}} \]
where $\tau$ is commonly denoted as the half-life.

@@

We choose to use windows with four times the half-life size (or $4\tau$), meaning we discard events whose contribution is less than 6\%. In fact, we set the half-life to be one quarter of the size of the window we want to emulate. For instance, if we want to compute an EMA aggregation over a period of a month, we set the half-life to be the number of tuples that correspond to one week worth of data.

@@

The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and the sliding window aggregations. In other words, we:

@@

Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase. To that end, we make \textit{S} samples of transactions, each with the same tuple-based size of the sliding window we will use in streaming. Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.

@@

For each feature, we end up with \textit{S} distance values. These values are distance measurements between random samples of data and the reference period. Hence, we claim we now know the distribution of expected distance values. Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if below a certain threshold. Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the sliding window histogram. This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.

It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself. For instance, given that JSD values varie between 0 and 1, why not just set to alert if the measured JSD value is 0.7? The reason is that a divergence value may fall on different percentiles for different features. In other words, the same \textit{i-th} percentile will correspond to a different divergence value for different features. Consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.

@@

Let's assume Figure \ref{fig:skewed-left-distro} 's histogram encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}. Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results. For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}. Instead what we aim to do is know this distribution so that we can define our threshold as a percentile and not a hard-coded distance constant. For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.

@@

\subsubsection*{Batch phase artifacts}
By the end of the batch phase, we should have, for each feature: A reference histogram, a list of distance values that represent its distribution and the last sample's histogram to be used as a burn-in period for the sliding window or target histogram.

@@

In order to obtain alerts, we periodically perform a multi-feature change detection test. The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the sliding window histogram of the same feature. This distance value is then tested against the known distribution of distance values for feature \textit{x}. In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile --- \textit{e.g.} the 99th-percentile. When we perform the test, for each feature \textit{x}, we are computing the percentile of the distance measured between the reference and sliding window histogram aggregations. The probability value or \textit{p-value} of a distance value is

@@

For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01. Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:

\[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]

or

\begin{equation}
    \label{eq:alert-test}
    \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha
\end{equation}

@@

\subsubsection*{Holm-Bonferroni Correction} \label{sec:holmbonferroni}
The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature. This correction is done by dividing the threshold $\alpha$ by a correction factor, as described below.

First, we compute the \textit{p-value} of each feature. We then order our features by \textit{p-value}, in ascending order. Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if

\[  \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha / (\textit{m} - \textit{k}) \]

with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature. We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.

@@

\subsubsection*{Optimal number of samples}
In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase. But what is the optimal number of samples to make? How can we mathematically derive a formula for the number of samples to make and avoid \textit{ad hoc} user-defined constants?

When checking for alerts we are performing one test for each feature, hence we performing \textit{n\textsubscript{features}} tests. We apply a multiple test correction on the tests performed and want to set the Family Wise Error Rate \textit{(FWER)} to $\alpha$. In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.

@@

\textcolor{red}{SEE MODEL MONITORING PAPER} We can then show that the optimal number of samples \textit{S} to make is

\begin{equation}
    \label{eq:optimal-n-samples}
    \textit{S} = \frac{\log\gamma}{log(1 - \frac{\alpha}{n\textsubscript{features}})}
\end{equation}

>>>>

\section{State of the art outlier detection methods}
Our first theoretical approach was to use one of the outlier detection methods presented in Section \ref{sec:outliers}. For that, we reviewed subsequence and point outlier detection methods. Using point outlier detection methods, the main idea is to keep track of the rate of point outliers detected. For instance, keep track of how many outliers were alerted in the past hour. The decision to alert for a data pattern shift of the system would be made when comparing this rate of point outliers detected in the past hour with a user-defined threshold $\alpha$. On the other hand, using subsequence outlier detection methods would allow us to report the whole sliding window as an outlier when compared to the reference window.

@@

In Section \ref{sec:sota-swag-algs} we reviewed algorithms to compute sliding window aggregations and in Sections \ref{sec:pds} and \ref{sec:sliding-pds} we discussed special aggregators and respective sliding window implementations.

@@

In Figure \ref{fig:approach2-alert-state}, after further sliding steps of our sliding window, we obtain \textit{Window 2}. In \textit{Window 2}, the mean and standard deviation aggregation values change from 5 and 1 to 12 and 6, respectively. In this case, we would measure the difference between the sliding window aggregations and the reference window ones and if above a certain threshold raise an alert. For example, if we used a threshold of $t = 6$, we would raise an alert, as at least one sliding aggregation, in this case the mean, would differ of at least $t$ when compared to the reference value.

@@

Another challenge when using this approach is defining the set of aggregations to compute at feature/variable level that represent the system state well enough for comparison of reference and streaming periods and change detection. To illustrate this problem, consider both time-series show in Figure \ref{fig:approach2-timeseries}. Time-series 1 and 2 have the same maximum, minimum, mean and standard deviation values for the time-based window between \textit{t\textsubscript{1}} and \textit{t\textsubscript{8}}. Hence, if our set of aggregations chosen were the maximum, minimum, mean and standard deviation, we would not differentiate between these two different time-series.

@@

In Section \ref{sec:stream-superset}, we defined a data stream as a continuous collection of events. Each event is timestamped and contains multiple fields. In the context of this Thesis, each of the event's fields are named features. Our goal is to detect data pattern shifts for each feature in an streaming fashion. Such shifts will be measured between a static in time or reference period versus the observed reality during the streaming phase. Figure \ref{fig:timelines} shows the separated reference and streaming timelines.

@@

To detect data pattern shifts, we do not need to directly compare the entire window contents of the reference and sliding windows. This approach would require storing all window contents hence making the memory consumption grow linearly with sliding window and/or reference period size. Instead, we want to use a  \textit{(a)} fixed cost in memory aggregation that can be \textit{(b)} incrementally maintained during the streaming phase. Additionally, we want a streaming aggregation that \textit{(c)} reliably encodes the distribution of each feature.

@@

To implement the proposed methodology, we devised a two-phase method. The first phase relies on a batch analysis that, for each feature, builds a reference histogram and computes the distribution of divergence values (details presented in Section \ref{sec:batch-phase}). The second phase (detailed in Section \ref{sec:stream-phase}) focuses on efficiently updating the sliding window histogram for each feature and periodically performing a divergence test between the static reference aggregation and the sliding one, alerting for changes if need be. The proposed method currently works only for numerical features, that is, only for event fields or variables that are of numeric type, but can be extended to other types.

\subsection{Exponential Moving Average based histogram} \label{sec:ema-hist}
As stated, we compute histogram aggregations for the reference and sliding windows. The computation of the reference histogram is done once in batch. Since the reference data is fixed and does not change over time, computing this histogram in batch is more efficient and easier to implement in a distributed system like Spark. In addition, a batch computation also allows us to compute an exact reference histogram. However, the sliding window histogram is an EMA-like histogram to ensure that we are able to encode the feature distribution with constant space complexity and maintain it in real-time.

In Section \ref{sec:emas} we discussed Exponential Moving Averages (EMAs) and how they compute sliding window aggregations without actually storing the window contents. Definition \ref{def:tuple-ema} shows how to compute an EMA using its recursive formula. An EMA state or value depends only on the previous one multiplied by an exponential decay $\lambda$. Recall that this decay is given by Equation \ref{eq:ema-decay}:
\[ \lambda = 2^{- \dfrac{1}{\tau}} \]
where $\tau$ is commonly denoted as the half-life.

@@

We choose to use windows with four times the half-life size (or $4\tau$), meaning we discard events whose contribution is less than 6\%. In fact, we set the half-life to be one quarter of the size of the window we want to emulate. For instance, if we want to monitor a period extending up to one month (at an error level of about 6\%) we setr the half-life to be the number of tuples that correspond to one week worth of data. This will give more weight to the most recent week but it will include contributions decaying to 6\% in the fourth oldest week.

@@

The goal of the batch phase will be to build our reference aggregation for each feature and get to know the distribution of divergence or distance values between the reference and EMA-like aggregations. In other words, we:

@@

Secondly, we aim to build the histogram that encodes the distribution of expected distance values, to later on threshold the observed distance values during the online phase. To that end, we make $S$ samples of transactions (in Section \ref{sec:nsamples} we discuss the optimal number $S$ of samples to make), each with the same tuple-based size of the sliding window we will use in streaming. Each sample is a contiguous block of transactions, thus preserving the order of transactions and the time-dependency property of a time-series.

@@

For each feature, we end up with \textit{S} distance values. These $S$ measurements are distance measurements between random samples of data and the reference period. Hence, we claim we now know the distribution of expected distance values. Given a new distance value measured between another random window and the reference window, we can compute its probability and produce alerts if the probability is below a certain threshold. Note that the histograms for each sample are EMA-like histograms (as detailed in Section \ref{sec:ema-hist}) just like the sliding window histogram. This ensures a certain degree of fidelity in the test we make in streaming because we measure the divergence between a random sample of data (our streaming histogram or any of the sample's histogram) and the reference one.

It is important to understand why we want to know the distribution of divergence values instead of just thresholding the divergence value itself. For instance, given that JSD values vary between 0 and 1, why not just set to alert if the measured JSD value is 0.7? The reason is that a divergence value may fall on different percentiles for different features. In other words, the same \textit{i-th} percentile will correspond to a different divergence value for different features. Consider the following left and right-skewed distance distributions represented by Figures \ref{fig:skewed-left-distro} and \ref{fig:skewed-right-distro}, respectively.

@@

Let's assume that the histogram from Figure \ref{fig:skewed-left-distro} encodes the distribution of expected distance values for a feature \textit{x1} and Figure \ref{fig:skewed-right-distro} does the same but for a feature \textit{x2}. Setting a user-defined threshold of \textit{$\alpha$=0.6} for both features would yield very different results. For feature \textit{x1} we would be reporting distance values above the 100th-percentile which would not be the case for feature \textit{x2}. Instead we define our threshold as a percentile of the distribution and not as a hard-coded distance constant. For instance, in this case, we would define the threshold to be the \textit{i}-th percentile, which would correspond to a threshold value of 0.4 for feature \textit{x1} and 0.6 for feature \textit{x2}.

@@

\subsubsection*{Batch phase artifacts}
By the end of the batch phase, we should have, for each feature: 
\begin{itemize}
    \item a reference histogram
    \item a list of distance values that represent its distribution
    \item the last sample's histogram, to be used as burn-in period for the sliding window histogram
\end{itemize}

@@

In order to obtain alerts, we periodically perform a multi-feature change detection test. The test is done by first computing the distance between the reference histogram for a feature \textit{x} and the sliding window histogram of the same feature. This distance value is then tested against the known distribution of distance values for feature \textit{x}. In other words, we test to see if the percentile of the computed distance is in the distribution of distance values and alert if above a certain percentile, \textit{e.g.}, the 99th-percentile. When we perform the test, for each feature \textit{x}, we compute the percentile of the distance measured between the reference and sliding window histogram aggregations. The probability value or \textit{p-value} of a distance value is:

@@

For example, if we define alerts to have 1\% or less probability to happen, it means the system will raise an alert for distance values above the 99th-percentile or \textit{p-value} = 0.01. Generally speaking, we alert a feature \textit{x} as divergent at a given timestamp if for a given probability threshold $\alpha$:
\[ \textit{percentile(distance\textsubscript{x})} > 1 - \alpha \]
or:
\begin{equation}
    \label{eq:alert-test}
    \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha
\end{equation}

@@

\subsubsection*{Holm-Bonferroni Correction} \label{sec:holmbonferroni}
The Holm-Bonferroni correction takes Equation \ref{eq:alert-test} and corrects the right-side threshold $\alpha$ for each feature. This correction is done by dividing the threshold $\alpha$ by a correction factor, as described next.

First, we compute the \textit{p-value} of each feature. We then order our features by \textit{p-value}, in ascending order. Next, starting from the first feature, the one with the smallest \textit{p-value}, and proceeding until the end of the list of features, we check if:
\[  \textit{p-value\textsubscript{distance\textsubscript{x}}} < \alpha / (\textit{m} - \textit{k}) \]
with $\alpha$ as the probability to reject the null hypothesis, \textit{m} as the number of hypotheses to test, in our case the number of features and corresponding \textit{p-values} and \textit{k} as the 0-based index of the current feature. We stop at the first feature that fails this test, and every feature processed up until this one is considered to be a divergent feature and alerts are raised.

@@

\subsubsection*{Optimal number of samples} \label{sec:nsamples}
In Section \ref{sec:sampling-batch} we stated we made \textit{S} samples in the batch phase. But what is the optimal number of samples to make? How can we mathematically derive a formula for the number of samples to make and avoid \textit{ad hoc} user-defined constants?

When checking for alerts we are performing one test for each feature, hence we perform \textit{n\textsubscript{features}} tests. We apply a multiple test correction on the tests performed and want to set the Family Wise Error Rate \textit{(FWER)} to $\alpha$. In other words, we want to alert deviations for the full set of tests globally that have probability $\alpha$ < 1\%, for example.

@@

\textcolor{red}{SEE MODEL MONITORING PAPER} We can then show that the minimum number of samples $S$ needed to satisfy these probability requirements is:
\begin{equation}
    \label{eq:optimal-n-samples}
    \textit{S} = \frac{\log\gamma}{log(1 - \frac{\alpha}{n_{features}})}
\end{equation}
